Endogenous Macrodynamics in Algorithmic Recourse

### 2022-11-08 paper/paper.tex
revision

A framework for Counterfactual Explanations was first proposed in 2017 by Wachter et al. \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} and has served as the baseline for most methodologies that have been proposed since then. Let \(M: \mathcal{X} \mapsto \mathcal{Y}\) denote some pre-trained model that maps from inputs \(X \in \mathcal{X}\) to outputs \(Y \in \mathcal{Y}\). Then we are interested in minimizing the cost\footnote{Equivalently, others have referred to this quantity as \emph{complexity} or simply \emph{distance}} \(C=\text{cost}(x^\prime)\) incurred by individual \(x\) when moving to a counterfactual state \(x^\prime\) such that the predicted outcome \(M(x^\prime)\) corresponds to some target outcome \(y^*\):

@@

We have chosen to focus on gradient-based counterfactual search for two reasons: firstly, they can be seen as direct descendants of our baseline method (Wachter); secondly, gradient-based search is particularly well-suited for differentiable black-box models like deep neural networks, which we focus on in this work. In particular, we include the following generators in our simulation experiments below: \textbf{REVISE} \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}}, \textbf{CLUE} \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}, \textbf{DiCE} \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and a greedy approach that relies on probabilistic models \protect\hyperlink{ref-schut2021generating}{{[}6{]}}. Our motivation for including these different generators in our analysis, is that they all offer slightly different approaches to generate meaningful counterfactuals for differentiable black-box models. We hypothesize that generating more \textbf{meaningful} counterfactuals should mitigate the endogenous dynamics illustrated in Figure \ref{fig:poc} in Section \ref{intro}. This intuition stems from the underlying idea that more meaningful counterfactuals are generated by the same or at least a very similar data generating process as the training data. All else equal, counterfactuals that fulfill this basic requirement should be less prone to trigger domain and model shifts.

@@

Algorithm \ref{algo-experiment} summarizes the proposed simulation experiment for a given dataset \(\mathcal{D}\), model \(M\) and generator \(G\), but naturally we are interested in comparing simulation outcomes for different sources of data, models and generators. The framework we have built facilitates this, making use of multi-threading in order to speed up computations. Holding the initial model and dataset constant, the experiments are run for all generators, since our primary concern is to benchmark different recourse methods. To ensure that each generator is faced with the exact same initial conditions in each round \(t\), the candidate batch of individuals from the non-target class is randomly drawn from the intersection of all individuals in the non-target class across all experiments \(\left\{\textsc{Experiment}(M,\mathcal{D},G)\right\}_{j=1}^J\) where \(J\) is the total number of generators.

@@

This section presents the exact ingredients and parameter choices describing the simulation experiments we ran to produce the findings presented in the next section (\ref{empirical-2}). For convenience, we use Algorithm \ref{algo-experiment} as a template to guide us through this section. A few high-level details upfront: each experiment is run for a total of \(T=50\) rounds, where in each round we provide recourse to five percent of all individuals in the non-target class, so \(B_t=0.05 * N_t^{\mathcal{D}_0}\)\footnote{As mentioned in the previous section, we end up providing recourse to a total of \(\approx50\%\) by the end of round \(T=50\).}. All classifiers and generative models are retrained for 10 epochs in each round \(t\) of the experiment. Rather than retraining models from scratch, we initialize all parameters at their previous levels (\(t-1\)) and compute backpropagate for 10 epochs using the new training data as inputs into the existing model. Evaluation metrics are computed and stored every 10 rounds. To account for noise, each individual experiment is repeat five times.\footnote{In the current implementation, we use the train-test split each time to only account for stochasticity associated with randomly selecting individuals for recourse. An interesting alternative may be to also perform data splitting each time, thereby adding an additional layer of randomness.}

@@

In this work we have focused on popular counterfactual generators that do not incorporate any causal knowledge. The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome. The implementation of such changes is typically described as \textbf{gaming} \protect\hyperlink{ref-miller2020strategic}{{[}36{]}}. Preventing such gaming may help to avoid the macrodynamics we have pointed to in this work. Future work would therefore likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al. \protect\hyperlink{ref-karimi2021algorithmic}{{[}13{]}}.

@@

For reasons stated earlier we have limited our analysis to differentiable linear and non-linear classifiers, in particular logistic regression and deep neural networks. While these sorts of classifiers have also typically been analyzed in the existing literature on Counterfactual Explanations and Algorithmic Recourse, they represent only a subset of popular machine learning models employed in practice. Despite the success and popularity of deep learning in the context of high-dimensional data such as image, audio and video, empirical evidence suggests that other models such as boosted decision trees may have an edge when it comes to lower-dimensional tabular datasets, such as the ones considered here (\protect\hyperlink{ref-borisov2021deep}{{[}37{]}}, \protect\hyperlink{ref-grinsztajn2022why}{{[}38{]}}).

@@

\leavevmode\vadjust pre{\hypertarget{ref-borisov2021deep}{}}%
\CSLLeftMargin{{[}37{]} }%
\CSLRightInline{V. Borisov, T. Leemann, K. Seßler, J. Haug, M. Pawelczyk, and G. Kasneci, {``Deep neural networks and tabular data: {A} survey,''} 2021. Available: \url{https://arxiv.org/abs/2110.01889}}

\leavevmode\vadjust pre{\hypertarget{ref-grinsztajn2022why}{}}%
\CSLLeftMargin{{[}38{]} }%
\CSLRightInline{L. Grinsztajn, E. Oyallon, and G. Varoquaux, {``Why do tree-based models still outperform deep learning on tabular data?''} 2022. Available: \url{https://arxiv.org/abs/2207.08815}}

>>>>

A framework for Counterfactual Explanations was first proposed in 2017 by Wachter et al. \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} and has served as the baseline for most methodologies that have been proposed since then. Let \(M: \mathcal{X} \mapsto \mathcal{Y}\) denote some pre-trained model that maps from inputs \(X \in \mathcal{X}\) to outputs \(Y \in \mathcal{Y}\). Then we are interested in minimizing the cost\footnote{Equivalently, others have referred to this quantity as \emph{complexity} or simply \emph{distance}.} \(C=\text{cost}(x^\prime)\) incurred by individual \(x\) when moving to a counterfactual state \(x^\prime\) such that the predicted outcome \(M(x^\prime)\) corresponds to some target outcome \(y^*\):

@@

We have chosen to focus on gradient-based counterfactual search for two reasons: firstly, they can be seen as direct descendants of our baseline method (Wachter); secondly, gradient-based search is particularly well-suited for differentiable black-box models like deep neural networks, which we focus on in this work. In particular, we include the following generators in our simulation experiments below: \textbf{REVISE} \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}}, \textbf{CLUE} \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}, \textbf{DiCE} \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and a greedy approach that relies on probabilistic models \protect\hyperlink{ref-schut2021generating}{{[}6{]}}. Our motivation for including these different generators in our analysis, is that they all offer slightly different approaches to generate meaningful counterfactuals for differentiable black-box models. We hypothesize that generating more \textbf{meaningful} counterfactuals should mitigate the endogenous dynamics illustrated in Figure \ref{fig:poc} in Section \ref{intro}. This intuition stems from the underlying idea that more meaningful counterfactuals are generated by the same or at least a very similar data generating process as the training data. All else equal, counterfactuals that fulfill this basic requirement should be less prone to trigger shifts.

@@

Algorithm \ref{algo-experiment} summarizes the proposed simulation experiment for a given dataset \(\mathcal{D}\), model \(M\) and generator \(G\), but naturally we are interested in comparing simulation outcomes for different sources of data, models and generators. The framework we have built facilitates this, making use of multi-threading in order to speed up computations. Holding the initial model and dataset constant, the experiments are run for all generators, since our primary concern is to benchmark different recourse methods. To ensure that each generator is faced with the same initial conditions in each round \(t\), the candidate batch of individuals from the non-target class is randomly drawn from the intersection of all non-target class individuals across all experiments \(\left\{\textsc{Experiment}(M,\mathcal{D},G)\right\}_{j=1}^J\) where \(J\) is the total number of generators.

@@

This section presents the exact ingredients and parameter choices describing the simulation experiments we ran to produce the findings presented in the next section (\ref{empirical-2}). For convenience, we use Algorithm \ref{algo-experiment} as a template to guide us through this section. A few high-level details upfront: each experiment is run for a total of \(T=50\) rounds, where in each round we provide recourse to five percent of all individuals in the non-target class, so \(B_t=0.05 * N_t^{\mathcal{D}_0}\)\footnote{As mentioned in the previous section, we end up providing recourse to a total of \(\approx50\%\) by the end of round \(T=50\).}. All classifiers and generative models are retrained for 10 epochs in each round \(t\) of the experiment. Rather than retraining models from scratch, we initialize all parameters at their previous levels (\(t-1\)) and compute backpropagate for 10 epochs using the new training data as inputs into the existing model. Evaluation metrics are computed and stored every 10 rounds. To account for noise, each individual experiment is repeated five times.\footnote{In the current implementation, we use the train-test split each time to only account for stochasticity associated with randomly selecting individuals for recourse. An interesting alternative may be to also perform data splitting each time, thereby adding an additional layer of randomness.}

@@

In this work we have focused on popular counterfactual generators that do not incorporate any causal knowledge. The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome. The implementation of such changes is typically described as \textbf{gaming} \protect\hyperlink{ref-miller2020strategic}{{[}36{]}}, although they need not be driven by adversarial intentions: in Example \ref{exm:student}, student applicants may dutifully focus on acquiring credentials that help them be admitted to university, but ultimately not improve their chances of success at completing their degree \protect\hyperlink{ref-barocas2017fairness}{{[}37{]}}. Preventing such actions may help to avoid the macrodynamics we have pointed to in this work. Future work would therefore likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al. \protect\hyperlink{ref-karimi2021algorithmic}{{[}13{]}}.

@@

For reasons stated earlier we have limited our analysis to differentiable linear and non-linear classifiers, in particular logistic regression and deep neural networks. While these sorts of classifiers have also typically been analyzed in the existing literature on Counterfactual Explanations and Algorithmic Recourse, they represent only a subset of popular machine learning models employed in practice. Despite the success and popularity of deep learning in the context of high-dimensional data such as image, audio and video, empirical evidence suggests that other models such as boosted decision trees may have an edge when it comes to lower-dimensional tabular datasets, such as the ones considered here (\protect\hyperlink{ref-borisov2021deep}{{[}38{]}}, \protect\hyperlink{ref-grinsztajn2022why}{{[}39{]}}).

@@

\leavevmode\vadjust pre{\hypertarget{ref-barocas2017fairness}{}}%
\CSLLeftMargin{{[}37{]} }%
\CSLRightInline{S. Barocas, M. Hardt, and A. Narayanan, {``Fairness in machine learning,''} \emph{Nips tutorial}, vol. 1, p. 2, 2017.}

\leavevmode\vadjust pre{\hypertarget{ref-borisov2021deep}{}}%
\CSLLeftMargin{{[}38{]} }%
\CSLRightInline{V. Borisov, T. Leemann, K. Seßler, J. Haug, M. Pawelczyk, and G. Kasneci, {``Deep neural networks and tabular data: {A} survey,''} 2021. Available: \url{https://arxiv.org/abs/2110.01889}}

\leavevmode\vadjust pre{\hypertarget{ref-grinsztajn2022why}{}}%
\CSLLeftMargin{{[}39{]} }%
\CSLRightInline{L. Grinsztajn, E. Oyallon, and G. Varoquaux, {``Why do tree-based models still outperform deep learning on tabular data?''} 2022. Available: \url{https://arxiv.org/abs/2207.08815}}

### 2022-11-08 paper/sections/empirical.rmd
revision

This section presents the exact ingredients and parameter choices describing the simulation experiments we ran to produce the findings presented in the next section (\@ref(empirical-2)). For convenience, we use Algorithm \ref{algo-experiment} as a template to guide us through this section. A few high-level details upfront: each experiment is run for a total of $T=50$ rounds, where in each round we provide recourse to five percent of all individuals in the non-target class, so $B_t=0.05 * N_t^{\mathcal{D}_0}$^[As mentioned in the previous section, we end up providing recourse to a total of $\approx50\%$ by the end of round $T=50$.]. All classifiers and generative models are retrained for 10 epochs in each round $t$ of the experiment. Rather than retraining models from scratch, we initialize all parameters at their previous levels ($t-1$) and compute backpropagate for 10 epochs using the new training data as inputs into the existing model. Evaluation metrics are computed and stored every 10 rounds. To account for noise, each individual experiment is repeat five times.^[In the current implementation, we use the train-test split each time to only account for stochasticity associated with randomly selecting individuals for recourse. An interesting alternative may be to also perform data splitting each time, thereby adding an additional layer of randomness.]

>>>>

This section presents the exact ingredients and parameter choices describing the simulation experiments we ran to produce the findings presented in the next section (\@ref(empirical-2)). For convenience, we use Algorithm \ref{algo-experiment} as a template to guide us through this section. A few high-level details upfront: each experiment is run for a total of $T=50$ rounds, where in each round we provide recourse to five percent of all individuals in the non-target class, so $B_t=0.05 * N_t^{\mathcal{D}_0}$^[As mentioned in the previous section, we end up providing recourse to a total of $\approx50\%$ by the end of round $T=50$.]. All classifiers and generative models are retrained for 10 epochs in each round $t$ of the experiment. Rather than retraining models from scratch, we initialize all parameters at their previous levels ($t-1$) and compute backpropagate for 10 epochs using the new training data as inputs into the existing model. Evaluation metrics are computed and stored every 10 rounds. To account for noise, each individual experiment is repeated five times.^[In the current implementation, we use the train-test split each time to only account for stochasticity associated with randomly selecting individuals for recourse. An interesting alternative may be to also perform data splitting each time, thereby adding an additional layer of randomness.]

### 2022-11-08 paper/sections/limitations.rmd
revision

In this work we have focused on popular counterfactual generators that do not incorporate any causal knowledge. The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome. The implementation of such changes is typically described as **gaming** @miller2020strategic. Preventing such gaming may help to avoid the macrodynamics we have pointed to in this work. Future work would therefore likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al. @karimi2021algorithmic.

>>>>

In this work we have focused on popular counterfactual generators that do not incorporate any causal knowledge. The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome. The implementation of such changes is typically described as **gaming** @miller2020strategic, although they need not be driven by adversarial intentions: in Example \@ref(exm:student), student applicants may dutifully focus on acquiring credentials that help them be admitted to university, but ultimately not improve their chances of success at completing their degree @barocas2017fairness. Preventing such actions may help to avoid the macrodynamics we have pointed to in this work. Future work would therefore likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al. @karimi2021algorithmic.

### 2022-11-08 paper/sections/methodology_2.rmd
revision

Algorithm \ref{algo-experiment} summarizes the proposed simulation experiment for a given dataset $\mathcal{D}$, model $M$ and generator $G$, but naturally we are interested in comparing simulation outcomes for different sources of data, models and generators. The framework we have built facilitates this, making use of multi-threading in order to speed up computations. Holding the initial model and dataset constant, the experiments are run for all generators, since our primary concern is to benchmark different recourse methods. To ensure that each generator is faced with the exact same initial conditions in each round $t$, the candidate batch of individuals from the non-target class is randomly drawn from the intersection of all individuals in the non-target class across all experiments $\left\{\textsc{Experiment}(M,\mathcal{D},G)\right\}_{j=1}^J$ where $J$ is the total number of generators.

>>>>

Algorithm \ref{algo-experiment} summarizes the proposed simulation experiment for a given dataset $\mathcal{D}$, model $M$ and generator $G$, but naturally we are interested in comparing simulation outcomes for different sources of data, models and generators. The framework we have built facilitates this, making use of multi-threading in order to speed up computations. Holding the initial model and dataset constant, the experiments are run for all generators, since our primary concern is to benchmark different recourse methods. To ensure that each generator is faced with the same initial conditions in each round $t$, the candidate batch of individuals from the non-target class is randomly drawn from the intersection of all non-target class individuals across all experiments $\left\{\textsc{Experiment}(M,\mathcal{D},G)\right\}_{j=1}^J$ where $J$ is the total number of generators.

### 2022-11-08 paper/sections/related.rmd
revision

A framework for Counterfactual Explanations was first proposed in 2017 by Wachter et al. @wachter2017counterfactual and has served as the baseline for most methodologies that have been proposed since then. Let $M: \mathcal{X} \mapsto \mathcal{Y}$ denote some pre-trained model that maps from inputs $X \in \mathcal{X}$ to outputs $Y \in \mathcal{Y}$. Then we are interested in minimizing the cost^[Equivalently, others have referred to this quantity as *complexity* or simply *distance*] $C=\text{cost}(x^\prime)$ incurred by individual $x$ when moving to a counterfactual state $x^\prime$ such that the predicted outcome $M(x^\prime)$ corresponds to some target outcome $y^*$:

>>>>

A framework for Counterfactual Explanations was first proposed in 2017 by Wachter et al. @wachter2017counterfactual and has served as the baseline for most methodologies that have been proposed since then. Let $M: \mathcal{X} \mapsto \mathcal{Y}$ denote some pre-trained model that maps from inputs $X \in \mathcal{X}$ to outputs $Y \in \mathcal{Y}$. Then we are interested in minimizing the cost^[Equivalently, others have referred to this quantity as *complexity* or simply *distance*.] $C=\text{cost}(x^\prime)$ incurred by individual $x$ when moving to a counterfactual state $x^\prime$ such that the predicted outcome $M(x^\prime)$ corresponds to some target outcome $y^*$:
